The Castro AMR Simulation Code: Current and Future Developments

### 2020-02-07 astronum_2019_Castro/astronum-castro-2019.tex
Fix typo: updated -> update
fix typo
Fix typo: summaries -> summarizes

In the SDC formalism~\cite{dutt:2000,minion:2003}, we start by writing our system in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc} + \Rbs{\Uc} \right ] dt
\end{equation}
We next introduce the idea of iteratively approximating the solution.
We'll use the superscript $k$ to denote the current iteration and
$k+1$ for the new iteration.  We rewrite the integral as:
\begin{align}
\Uc^{n+1,(k+1)} = \Uc^{n,(k+1)} +
   &\int_{t^n}^{t^{n+1}} \left \{ \left [ \Advs{\Uc}^{(k+1)} - \Advs{\Uc}^{(k)} \right ]
   + \left [ \Rbs{\Uc}^{(k+1)} - \Rbs{\Uc}^{(k)} \right ] \right \} dt + \nonumber \\
  & \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc}^{(k)} + \Rbs{\Uc}^{(k)} \right ] dt
\end{align}
For the first integral, we approximate these terms using a first order
update.  For advection, we simply use the value at the starting time
(analogous to a forward Euler approximation) and for the reactions, we
treat them implicitly by using the value at the ending time (analogous
to a backward Euler approximation).  To gain temporal accuracy, we
further subdivide the time into $M+1$ nodes, and perform this update
over all time nodes for each iteration.  Finally, in \castro\ we use a finite-volume
discretization, so we'll write the state as $\avg{\Uc}$ to indicate that it is an average over the zone.
Denoting the time nodes with
the index $m$, the updated from one time node $m$ to $m+1$ appears as:
    \begin{align}
      \label{eq:sdc:general}
      \avg{\Uc}^{{m+1},(k+1)} = \avg{\Uc}^{m,(k+1)}
            &+ \delta t_m \left [ \avg{\Advs{\Uc}}^{m,(k+1)} - \avg{\Advs{\Uc}}^{m,(k)} \right ] \nonumber \\
            &+ \delta t_m \left [ \avg{\Rbs{\Uc}}^{{m+1},(k+1)} - \avg{\Rbs{\Uc}}^{{m+1},(k)} \right ] \nonumber \\
            &+ \int_{t^m}^{t^{m+1}} dt  \left (\avg{\Advs{\Uc}}^{(k)} + \avg{\Rbs{\Uc}}^{(k)}\right)
   \end{align}
where $\delta t_m = t^{m+1} - t^m$ is the time difference between nodes.
Here, $\avg{\Advs{\Uc}}^m$ is the
average advective term in a zone at time $t^m$, and
$\avg{\Rbs{\Uc}}^{m+1}$ the average of the reactive source at time
$t^{m+1}$.  The second superscript on each term, $(k)$ or $(k+1)$, is
the iteration counter.  This update is an implicit equation for the
new state, $\avg{\Uc}^{{m+1},(k+1)}$.  The last term in the update is
an integral over the sources constructed from the previous iteration's
values at each time node.  Each iteration of the SDC method increases
the temporal order of accuracy by one, up to the order of accuracy
with which the integral is constructed.

\castro\ implements both a second-order method (using a trapezoid rule
for the integral) and fourth-order method in space and time (using a
Simpsons rule for the integral and the spatial reconstruction of
\cite{mccorquodalecolella}).  We demonstrated that \castro\ achieves
the expected convergence on a wide variety of problems in
\cite{castro:sdc}.  Table~\ref{table:convergence} summaries the
convergence of the standard unsplit CTU PPM hydrodynamics algorithm
with Strang splitting and the second- and fourth-order SDC methods on
a reacting acoustic pulse test problem modeling the 3-$\alpha$ and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ reactions.  We see that
the Strang CTU method has trouble achieving second order accuracy,
while the second-order SDC method converges second order.  The
fourth-order SDC method converges almost fourth order overall.  Full
details of this problem and more extensive convergence results are
given in \cite{castro:sdc}.

>>>>

In the SDC formalism~\cite{dutt:2000,minion:2003}, we start by writing our system in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc} + \Rbs{\Uc} \right ] dt
\end{equation}
We next introduce the idea of iteratively approximating the solution.
We'll use the superscript $k$ to denote the current iteration and
$k+1$ for the new iteration.  We rewrite the integral as:
\begin{align}
\Uc^{n+1,(k+1)} = \Uc^{n,(k+1)} +
   &\int_{t^n}^{t^{n+1}} \left \{ \left [ \Advs{\Uc}^{(k+1)} - \Advs{\Uc}^{(k)} \right ]
   + \left [ \Rbs{\Uc}^{(k+1)} - \Rbs{\Uc}^{(k)} \right ] \right \} dt + \nonumber \\
  & \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc}^{(k)} + \Rbs{\Uc}^{(k)} \right ] dt
\end{align}
For the first integral, we approximate these terms using a first order
update.  For advection, we simply use the value at the starting time
(analogous to a forward Euler approximation) and for the reactions, we
treat them implicitly by using the value at the ending time (analogous
to a backward Euler approximation).  To gain temporal accuracy, we
further subdivide the time into $M+1$ nodes, and perform this update
over all time nodes for each iteration.  Finally, in \castro\ we use a finite-volume
discretization, so we'll write the state as $\avg{\Uc}$ to indicate that it is an average over the zone.
Denoting the time nodes with
the index $m$, the update from one time node $m$ to $m+1$ appears as:
    \begin{align}
      \label{eq:sdc:general}
      \avg{\Uc}^{{m+1},(k+1)} = \avg{\Uc}^{m,(k+1)}
            &+ \delta t_m \left [ \avg{\Advs{\Uc}}^{m,(k+1)} - \avg{\Advs{\Uc}}^{m,(k)} \right ] \nonumber \\
            &+ \delta t_m \left [ \avg{\Rbs{\Uc}}^{{m+1},(k+1)} - \avg{\Rbs{\Uc}}^{{m+1},(k)} \right ] \nonumber \\
            &+ \int_{t^m}^{t^{m+1}} dt  \left (\avg{\Advs{\Uc}}^{(k)} + \avg{\Rbs{\Uc}}^{(k)}\right)
   \end{align}
where $\delta t_m = t^{m+1} - t^m$ is the time difference between nodes.
Here, $\avg{\Advs{\Uc}}^m$ is the
average advective term in a zone at time $t^m$, and
$\avg{\Rbs{\Uc}}^{m+1}$ the average of the reactive source at time
$t^{m+1}$.  The second superscript on each term, $(k)$ or $(k+1)$, is
the iteration counter.  This update is an implicit equation for the
new state, $\avg{\Uc}^{{m+1},(k+1)}$.  The last term in the update is
an integral over the sources constructed from the previous iteration's
values at each time node.  Each iteration of the SDC method increases
the temporal order of accuracy by one, up to the order of accuracy
with which the integral is constructed.

\castro\ implements both a second-order method (using a trapezoid rule
for the integral) and fourth-order method in space and time (using a
Simpsons rule for the integral and the spatial reconstruction of
\cite{mccorquodalecolella}).  We demonstrated that \castro\ achieves
the expected convergence on a wide variety of problems in
\cite{castro:sdc}.  Table~\ref{table:convergence} summarizes the
convergence of the standard unsplit CTU PPM hydrodynamics algorithm
with Strang splitting and the second- and fourth-order SDC methods on
a reacting acoustic pulse test problem modeling the 3-$\alpha$ and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ reactions.  We see that
the Strang CTU method has trouble achieving second order accuracy,
while the second-order SDC method converges second order.  The
fourth-order SDC method converges almost fourth order overall.  Full
details of this problem and more extensive convergence results are
given in \cite{castro:sdc}.
